An assessment of the biodiversity - ecosystem function relationship in southern African woodlands

### 2020-06-29 manuscript/regional_befr_manuscript.tex
Updated with comments

\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear}. Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density, resource availability \rnew{, disturbance through fire}, and across vegetation type to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

@@

\textbf{Methods:} We used a network of \nplots{} savanna/woodland plots located across the southern African sub-continent in which each tree >10 cm diameter was measured \rnew{and identified to species level}. We used Structural Equation Modelling and path analysis to determine the relationship between tree species diversity and aboveground woody biomass, accounting for the interactive effects of resource availability \rnew{and disturbance by fire} along a gradient of stem density.

\textbf{Results:} A positive effect of tree species diversity on aboveground biomass was demonstrated, observed largely as an indirect effect of increasing woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability affects biomass in southern African woodlands largely indirectly via its effect on species diversity.

\textbf{Main conclusions:} The study underlines the close association between tree diversity, ecosystem structure and function of highly disturbed southern African savannas and woodlands. Our results demonstrate the importance of including environmental conditions and vegetation type in models to accurately describe a general relationship between biodiversity and ecosystem function at a regional level. Biodiversity loss predominantly by human actions in southern Africa may have detrimental outcomes for ecosystem function, particularly in species poor Baikiaea woodlands, which showed the strongest biodiversity - ecosystem function relationship.

@@

Scientific interest in the relationship between species diversity and ecosystem function springs from both an interest in the factors which structure ecological communities \citep{}, and a more applied interest in determining the effect of global biodiversity loss on ecosystem form and function \citep{}. Numerous studies have shown relationships between biodiversity and ecosystem function (e.g. \citealt{Liang2016, Hooper2012, Cardinale2009}). The strength and direction of these observed Biodiversity-Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small scale experimental studies predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}. This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported and intuitive relationship between biodiversity and ecosystem function invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater biodiversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.

Compared to \rchange{other forested}{forest} ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire \rchange{and in Africa notably by herbivory also}{and herbivory, with African savannas possessing large herbivores absent from other savannas} \citep{Sankaran2008, Levick2009}. \rnew{Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{}}. High levels of disturbance may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as apparent in dry woodland/savanna ecosystems. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem function\rnew{s} \citep{Lasky2014, Tobner2016}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations \rchange{in environments which are more hostile to growth}{under limiting environmental conditions such as low water availability \citep{} or high maximum temperature \citep{}}. Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, \rnew{as competition dminishes in ecosystems where environmental stress limits individual species reaching the bounds of their fundamental niche space.} This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation in order to characterise a generalisable BEFR.

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanism described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and its environmental determinants. Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species. Southern African savannas and woodlands (SAWs) are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely un-quantified consequences for ecosystem structure and function.

A small number of studies in SAWs, all of which were restricted in the spatial scope to a small region of miombo woodland, have found that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine scale studies concur with similar studies in other biomes \citep{}. Studies of the BEFR often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Due to the highly variable environmental conditions within which SAWs occur \citep{Frost1996}, \rnew{with wide variation in precipitation, diurnal and annual temperature range} and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.

@@

\citet{Solbrig1996} writes that SAWs possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species. This structural diversity may be one mechanism through which tree species diversity influences woody biomass. \citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production. Occupation of multiple canopy layers allows a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition. Instead, disturbance history may override the effects of tree species diversity on structural diversity nullifying the effects of tree species diversity on structural diversity.

High levels of disturbance in SAWs may moderate the observable BEFR through its effect on ecosystem composition. Fire disturbance in forests has been linked to abundance dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}. Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree. Selection effects may therefore be more important in maximising ecosystem function in disturbance prone woodlands. If the regional species pool contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.

In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands (SAWs), using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}). We used aboveground woody biomass of trees and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type via clustering of plot level tree species composition, as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and assess the generality of our results. We used Structural Equation Modelling (SEM) and path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass. Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strenghth of the effect} of tree species diversity on woody biomass will increase with plot level stem stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.

@@

Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands (SAWs) \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates. Only plots with a stem density >10 stems ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.

@@

After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition. We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbour criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers. We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, thus removing \noutliers{} plots (\hyperref[appendixa]{Appendix A}).

@@

We considered only trees and shrubs in our calculations of above-ground woody biomass (AGB), including woody species such as palms and cycads which are functionally tree-like, but excluding lianas, which fill a different ecological niche \citep{Selaya2008}. Only stems >10 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses. Many plots in the dataset did not include data on stems <10 cm DBH. For those plots with stem measurements <10 cm DBH, those small stems only accounted for a median average of \percsmallagb{}\% of the plot level AGB.

All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\nstems{}} stems with measurements. A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual. For each stem we measured species, DBH and tree height to the top of the highest branch material. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from (Ryan C., unpublished) (\hyperref[appendixb]{Appendix B}).

AGB for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}. Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}. Wood density for species without species level estimates was estimated from the mean of their respective genus.

@@

Where $\rho$ is the species level mean wood density, $D$ is the DBH\textsubscript{e} at 1.3 m, and $H$ is the tree height.

@@

Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}. This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort among plots. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}. To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness. Abundance evenness allows for greater niche complementarity at small scales due to an increased spatial heterogeneity of functional traits. We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CV) and tree height (Height CV).

@@

Plots were assigned to vegetation type groups based on tree species composition. Groups were defined in \citet{Fayolle2018} in an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data. Group identification was conducted using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes. Plot data used in this study occurred in four vegetation type groups. See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters .

@@

\begin{figure}[H]
\centering
	\includegraphics[width=1.4\textwidth]{clust_map}
	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of miombo woodland vegetation according to \citet{White1987}. Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
	\label{clust_map}
\end{figure}
\end{landscape}

@@

Structural Equation Models (SEM) investigated the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is challenging in standard multiple regressions. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset. Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.

@@

Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (\hyperref[appendixc]{Appendix C}). Standardisation put each latent variable on the same scale, with a mean of zero and a standard deviation of one. Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}. Standardisation also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|. To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables were reversed by multiplying by -1. For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed for model fitting. Precipitation seasonality (PS), temperature seasonality (TS), and mean annual temperature (MAT) were also reversed in this way to account for the direction of their effect on moisture availability.

@@

In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity had a small positive direct effect on AGB (\strucdb{}), and indirectly via structural diversity (\strucdsb{}) (\autoref{struc_mod}). Tree species diversity had a positive correlation with stem density. Model fit was good with high factor loadings for all observed variables, all path coefficients were significant (p <0.01) (\autoref{struc_model_fit_clust_stats}). The R\textsuperscript{2} of AGB was \strucrsq{}. The strongest direct effect on AGB was from stem density (\strucsib{}).

>>>>

\textbf{Aim:} Positive correlations between tree \rchange{biodiversity}{species diversity} and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear}. Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, \rchange{with respect to}{while controlling for} gradients in stem density, resource availability \rnew{, disturbance through fire}, and across vegetation type to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

@@

\textbf{Methods:} We used a network of \nplots{} \rchange{savanna and woodland}{savanna/woodland tree} plots located across the southern African sub-continent \rold{in which each tree >10 cm diameter was measured and identified to species level}. We used Structural Equation Modelling and path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability \rnew{, disturbance by fire}, and stem density.

\textbf{Results:} \rnew{We found} A positive effect of tree species diversity on aboveground biomass \rold{was demonstrated}, observed largely as an indirect effect of increasing woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability affects biomass in southern African woodlands \rchange{largely}{mainly} indirectly, via its effect on species diversity.

\textbf{Main conclusions:} The study underlines the close association between tree diversity, ecosystem structure and \rnew{ecosystem} function of highly disturbed southern African savannas and woodlands. Our results demonstrate the importance of \rchange{including}{accounting for} environmental conditions and vegetation type in order to accurately model a general relationship between biodiversity and ecosystem function at a regional level. Biodiversity loss, predominantly \rchange{by}{through} human actions in southern Africa may have detrimental outcomes for ecosystem function, particularly in tree species-poor Baikiaea woodlands, which showed the strongest biodiversity - ecosystem function relationship.

@@

Scientific interest in the relationship between species diversity and ecosystem function springs from both an interest in the factors which structure ecological communities \citep{}, and a more applied interest in determining the effect of global biodiversity loss on ecosystem form and function \citep{}. Numerous studies have shown relationships between biodiversity and ecosystem function (e.g. \citealt{Liang2016, Hooper2012, Cardinale2009}). The strength and direction of these observed Biodiversity-Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small-scale experimental studies \rnew{conducted} predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}. This includes \rchange{basic processes of primary production}{processes} such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported \rold{and intuitive} relationship between biodiversity and ecosystem function invokes three main mechanisms \rchange{which drive}{to explain} the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.

Compared to \rchange{other forested}{forest} ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire \rchange{and in Africa notably by herbivory also}{and herbivory, with African savannas possessing large herbivores absent from other savannas} \citep{Sankaran2008, Levick2009}. \rnew{Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{}}. High levels of disturbance may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as \rchange{apparent}{important} in dry woodland/savanna ecosystems. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations \rchange{in environments which are more hostile to growth}{under limiting environmental conditions such as low water availability \citep{} or high maximum temperature \citep{}}. Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, \rnew{as competition dminishes in ecosystems where environmental stress limits individual species reaching the bounds of their fundamental niche space.} This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation \rchange{in order to characterise}{if we are to derive} a generalisable BEFR.

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanisms described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and its environmental determinants. Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages\rchange{, which}{.These} retain the common features of an open tree canopy and an understorey generally dominated by C4 \change{grass species}{grasses}. Southern African savannas and woodlands (SAWs) are highly diverse, thought to harbour \textapprox{}8500 plant species of which \rold{there are} >300 \rnew{are} \rchange{tree species}{trees} \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely unquantified consequences for ecosystem structure and function.

A small number of studies in SAWs, \rnew{all of which were restricted in the spatial scope to a small region of miombo woodland}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine scale studies concur with similar studies in other biomes \citep{}. Studies of the BEFR often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Due to the highly variable environmental conditions within which SAWs occur \citep{Frost1996}, \rnew{with wide variation in precipitation, diurnal and annual temperature range} and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.

@@

\citet{Solbrig1996} writes that SAWs possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species. This structural diversity may be one mechanism through which tree species diversity influences woody biomass. \citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species-rich neighbourhoods exhibited enhanced biomass production. \rchange{Occupation}{Occupancy} of multiple canopy layers allows a \rchange{more full}{fuller} canopy with \rold{a} greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas. Instead, the overriding importance of disturbance history may negate the effects of tree species diversity on structural diversity \citep{Grime2012}.

High levels of disturbance in SAWs may moderate the observable BEFR through its effect on ecosystem composition. Fire disturbance in forests has been linked to abundance-dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}. Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree. Selection effects may therefore be more important in maximising ecosystem function in disturbance prone woodlands. If the regional species pool contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.

In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands (SAWs), using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}). We used aboveground woody biomass of trees \rnew{as our metric of ecosystem function}, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type (via clustering of plot level tree species composition), as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and to assess the generality of our results. We used Structural Equation Modelling (SEM) \rchange{and}{with} path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass. Initially, we \rchange{made}{posited} three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strength of the effect} of tree species diversity on woody biomass will increase with plot level stem stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.

@@

Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands (SAWs) \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool. Only plots >0.1 hectares were used in analysis, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates. Only plots with a stem density >10 stems ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.

@@

\rold{After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition. We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbour criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers. We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, thus removing \noutliers{} plots (\hyperref[appendixa]{Appendix A}).}

@@

We considered only trees and shrubs in our calculations of above-ground woody biomass (AGB), including woody species such as palms and cycads which are functionally tree-like, but excluding lianas, which fill a different ecological niche \citep{Selaya2008}. Only stems >10 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses. Many plots in the dataset did not include data on stems <10 cm DBH. For those plots which contained stem measurements <10 cm DBH, small stems only accounted for a median of \percsmallagb{}\% of the plot level AGB.

All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\nstems{}} stems with measurements. A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual. For each stem, we \rchange{measured}{recorded} species, DBH and tree height to the top of the highest branch material. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from Ryan C., (unpublished), see  \citet{Godlee2020}.

AGB for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}:

@@

where $\rho$ is the species level mean wood density, $D$ is the DBH\textsubscript{e} at 1.3 m, and $H$ is the tree height. Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}. Wood density for species without species level estimates was estimated from the means of their respective \rchange{genus}{genera}.

@@

Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}. This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort among plots. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}. To measure tree species \rold{abundance} evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness. Abundance evenness allows for greater niche complementarity at small scales due to an increased spatial heterogeneity of functional traits. We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CV) and tree height (Height CV).

@@

Plots were assigned to vegetation type groups based on tree species composition. Groups were defined in \citet{Fayolle2018} in an Africa-wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data. Group identification was conducted using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes. Plot data used in this study occurred in four vegetation type groups. See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters .

@@

\begin{figure}[H]
\centering
	\includegraphics[width=1.4\textwidth]{clust_map}
	\caption{The locations of the \nplots{} plots used in this study, with respect to the distribution of miombo woodland vegetation according to \citet{White1987}. Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
	\label{clust_map}
\end{figure}
\end{landscape}

@@

\rnew{We used} Structural Equation Models (SEM) to investigate the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEMs is that they can explicitly model and partition variance \rnew{attributed} to indirect effects, which is challenging in standard multiple regressions. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having observable measures in our dataset. Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.

@@

Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (\hyperref[appendixa]{Appendix A}). Standardisation put each latent variable on the same scale, with a mean of zero and a standard deviation of one. Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}. Standardisation also controls for variables with different orders of magnitude, which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|. To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables were reversed by multiplying by -1. For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed for model fitting. Precipitation seasonality (PS), temperature seasonality (TS), and mean annual temperature (MAT) were also reversed in this way to account for the direction of their effect on moisture availability.

@@

In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity had a small positive direct effect on AGB (\strucdb{}), and indirectly via structural diversity (\strucdsb{}) (\autoref{struc_mod}). Tree species diversity had a positive correlation with stem density. Model fit was good with high factor loadings for all observed variables. All path coefficients were significant (p <0.01) (\autoref{struc_model_fit_clust_stats}). The R\textsuperscript{2} of AGB was \strucrsq{}. The strongest direct effect on AGB was from stem density (\strucsib{}).
